Adapting the Chumbley Score to match striae on Land Engraved Areas of bullets

### 2018-08-07 Response.Rmd

Because we are trying to facilitate timely publication of manuscripts submitted to the Journal of Forensic Sciences, your revised manuscript should be uploaded within 21 days of receipt of this email. If it is not possible for you to submit your revision in this time frame, we may have to consider your paper as a new submission.

Once again, thank you for submitting your manuscript to the Journal of Forensic Sciences and I look forward to receiving your revision.

## Reviewer #1

The article "Adapting the Chumbley Score to match striae on Land Engraved Areass (LEAs) of bullets" extends an existing comparison algorithm for striated marks to the analysis of bullet LEAs. The paper makes two primary contributions. First, they modify the deterministic work of Hadler & Morris (2017) which itself is an extension of Chumbley et al. The authors' contribution is a modification of Hadler & Morris to reduce the number of scenarios which 'fail' under the Hadler & Morris approach. Second, they evaluate the work on two sets of bullet LEA data and explore some variants of the algorithm parameters to tune it for bullets rather than screwdrivers. Overall the paper presents a reasonable project to extend a nice algorithm for striated toolmark comparison.

@@

Adaptation of the Chumbley el al algorithm from screwdrivers to bullets makes sense as both analyses involve striated marks. The authors here nicely point out that compared to screwdriver marks (as in the original Chumbley paper), LEAs are typically smaller both in length and width. LEAs also have a curved cross-section baseline which must be corrected.

The authors do a nice job covering previous work. I appreciate the summary tables 1 and 2. Table 1 is good as-is. Table 2 can be a bit misleading because we don't know what bullet datasets were used. Because these different papers analyzed different bullet data sets the underlying difficulty of matching impacts the performance. That is, bullets from very well marking firearms are 'easier' to match than those from firearms which mark poorly. So the FP and FN reported should be taken with a grain of salt. The authors should add a statement regarding different performance on different datasets.

I'm also confused how the FP rate for Hare CNMS is 6.26% whereas all other FP are 6.25% This seems to likely be a typo. 6.25% imply close to 10,000 comparisons which I don't think the datasets have. I could be wrong but I wanted to point out this potential typo that should be checked.

The authors describe the two methods of 'failure' for the method (Sec 2.3). I agree with the authors statement that this failure 'should' only happen for non-matches. It could also happen for matches if the optimal profiles are incorrect. That is, if there are two possible optimal match pairs (with very similar correlation scores), where the top scoring pair is incorrect but the second best scoring pair is correct. I'm not sure how often that happens. Overall, I would think that failing for non-matches would be ok. It's not clear that the described failures are the biggest cause of performance hits but improving the algorithm to handle these failures seems reasonable.

@@

The following sections list specific points that should be addressed before acceptance. They are grouped into clarification/grammar/preciseness, minor, and major revisions.

Some examples of clarification / grammar / preciseness issues:

- p3 "confocal microscopy allow * to measure 3D surfaces in a high-resolution digital form"  * should be "one"?
- p3 "Digitized images of 3D surfaces of form the data basis* of statistical analysis" Awkward phrasing.
- p3 "Faden et al and Chumbley et al have been analyzing screwdriver marks" perhaps change to "have analyzed" because the work is completed and published.
- p9 Seems to be a broken latex figure reference around line 28/29.
These are not all the issues of this type, we encourage the authors to review the manuscript carefully clean issues like the above.

Minor edits should be made to accuracy and clarity:

- p3 "Methods for matching marks for a variety of tools" should be changed to "Methods for matching striated marks for a variety of tools". This is an important point in that it helps make clear that the authors are studying and comparing striated toolmarks and not impressed toolmarks. I would also propose that the authors make this distinction earlier in the introduction.
- p3 "Analysis of these digitized markings require the use of statistical methods". Analysis does not *require* statistical methods. Useful information regarding common origin can be obtained from non-statistical methods. This being said, statistical methods are of course great and we should be working towards them. Authors could change to analysis "benefits from the use of statistical methods" or something similar.
- p14 how are the false negative and false positive rates computed?

@@

Major Changes Required before acceptance:

- p8 "This severe limitation in the amount of available data poses the main challenge in". It's unclear what 'this' refers to. Do the authors mean the number of bullets and LEAs in the dataset or do they mean the number of sample points per striation profile (due to the fact that LEAs are shorter than screwdriver profiles)? Both would seem to be limitations but just because the profile is shorter doesn't inherently mean that there are fewer or less informative striaes. You could have a shorter profile with more informative features.
- Eq 1 uses a variable i, which is defined above as having the values of 1 or 2. I therefore don't understand how i<0 as written in Eq 1. This should be improved. Possibly this is a sloppy reuse of variables?
- How are the original profiles (eg. Figure 2 Left) identified from the 3D surface (eg. Figure 1 Top). That is, does the user simply draw a line from one end of the LEA to the other? How is the original 3D surface resampled to the 1d profile? This is very important because if you 'cut' the 3D profile the wrong way (at the wrong angle) then you will dilate or compress the extracted profile. This needs a lot more description.
- Section 2.4 is not clearly written. I'm not confident that I understand their new method. This section should be rewritten to explain what the authors are doing. My guess is that rather than taking windows in integer number of steps away from the optimal position matching a left shift on one profile with a right shift on the other, the authors simply walk from the right-end to the left-end of one profile and from the left-end to the right-end of the other profile being careful to not compare same shift windows. Note that my description is not clear either as this is a tricky thing to put into words, but is something the authors need to do.
- The results presented on page 14 list the correct identification of "same-source and difference-source toolmarks of 93.5 to 94.1%". Does this mean the true positives are 93.5% and the true negatives are 94.1%? I don't think so because the next sentence lists the false negative rate of between 0.28 and 0.36. The authors need to explicitly list the true positive rate and the true negative rate. These must be separate because there are many more true negatives than true positives in any matching dataset. That means that an algorithm that always predicts "different source" would be correct most of the time. In fact, for the author's dataset the "always predict different" algorithm would be correct approximately 94% of the time (35 bullets where each has appx 2 matches). The lack of a true positive rate and true negative rate is is a major point that must be corrected. To be clear, the true positive rate, true negative rate, false positive rate, and false negative rate should all be reported for the first dataset, perhaps added to Table 3 or another table like Table 3.
- The true positive rate, true negative rate, false positive rate, and false negative rate should also be presented for the second dataset (Hamby 44 scans, Sec 3.5). The authors do present AUC and EER curves but these are less informative and can often be used to hide weaker performing models.
- The results should all be listed as either fractions or all as percent (which is preferred) and not mix them.
- p21 the authors state "this kind of assessment is only feasible in the setting of a large study, such as the one we presented".  I strongly disagree with the use of the word "large". The presented work is not a large study. Ten firearms were used in each of two studies (it's not clear if both datasets used the same 10 firearms). A large study would include at least 100 different firearms and even 100 is not likely to provide statistical confidence. This is a very important point. If the paper is allowed to state that the described work is a large study it will set a precedent that will be detrimental to the field.

@@

Overall, this paper presents a nice incremental improvement to a promising algorithmic approach. A number of major revisions are required prior to acceptance as described above. Note that these revisions all relate to the writing and require no additional experiments to be performed. They improve the clarity and preciseness of the manuscript and serve to address questions that readers will have.

## Reviewer #2

You may not fully appreciate and how critical their phase orientations used to identify fired bullets (LEAs).  This important consideration should have been an important  aspect of the experiment and not instead focus primarily only LEA comparisons.  While each area is in practice independently compared microscopically or by surface profile algorithms, consideration as to their orientation in their position in the barrel rifling, and correlations of LEAs cannot ignore the phase of positions.  Including the phase consideration of the fired bullet comparisons, the success of finding a match in a particular position between two bullets is significantly increased and demonstrated in their reference Chu, et al.  If you are to base independent single LEA scoring to test the method, you should make this clearer in the beginning of your paper and hypothesis, and not discuss it finally in the last sentence of the document body.
If you are going to compare method performance between chisel/tool stria to LEAs on bullets, you should have a detailed discussion as to the large differences in the toolmark widths that are in consideration.  The Chumbley scoring parameters may have been optimized for chisel widths but may not be optimal for the LEAs.

### 2018-08-07 doc/paper.Rmd
can't get the references to work, but the style file is closer to JFS specs

---
title: "Adaption of the Chumbley Score to matching of bullet striation marks"
authors:
- affiliation: Department of Statistics, Iowa State University 
  name: Ganesh Krishnan
  thanks: The authors gratefully acknowledge funding from CSAFE through cooperative agreement \# 70NANB15H176 between NIST and Iowa State University. 
- affiliation: Department of Statistics and CSAFE, Iowa State University 
  name: Heike Hofmann
biblio-style: apsr
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
    keep_tex: yes
    latex_engine: pdflatex
    template: template.tex
  html_document: default
blinded: 0
keywords:
- forensic science
- toolmark
- cross-correlation
- Mann-Whitney U statistic
- land engraved areas (LEAs)
- algorithm
- signatures 
- same-source problem
bibliography: bibliography
abstract: The same-source problem remains a major challenge in forensic toolmark and firearm examination. Technological advances in surface imaging allow measurements of 3D surfaces at previously unforeseen resolutions and enable digitized imaging. Here, we investigate the applicability of the Chumbley scoring method \citep{hadler}, developed for screwdriver markings, for same-source identification of bullet striae. We provide methods to identify parameters that minimize the error rates for matching of LEAs using Hamby datasets 44 and 252 measured by NIST and CSAFE. We suggest a remedial algorithm to alleviate the problem of failed tests in the method, increasing both the power of the test and reducing error rates. Type II error rates of the proposed method improve by more than one third (Type I error of 0.05) and are on average about 0.22. This puts the proposed method on similar footing as other single feature matching approaches in the literature. 
---

>>>>

---
title: "Adapting the Chumbley Score to match striae on Land Engraved Areas of bullets"
authors:
# - affiliation: Department of Statistics, Iowa State University 
- name: Ganesh Krishnan,
  degree: Master of Science (M.Sc.)
  # thank: The authors gratefully acknowledge funding from CSAFE through cooperative agreement \# 70NANB15H176 between NIST and Iowa State University. We would also like to thank the efforts of CSAFE and Roy J carver Microscopy lab in scanning the Hamby-44 set and providing the scans to us.
# - affiliation: Department of Statistics and CSAFE, Iowa State University 
- name: Prof. Heike Hofmann,
  degree: Doctor of Philosophy (Ph.D.)
affiliation: Department of Statistics, Iowa State University and CSAFE (Center for Statistics and Applications in Forensic Evidence)
funding: The authors gratefully acknowledge funding from CSAFE through cooperative agreement \# 70NANB15H176 between NIST and Iowa State University. 
thank: We would like to thank the efforts of CSAFE and Roy J Carver Microscopy lab in scanning the Hamby-44 set and providing the scans to us. We would also like to thank NIST for providing the Hamby-44 bullets to us for scanning.
presentedat:  This work was presented in the form of a poster in the AFTE 2018 49th Annual Training Seminar, held from June 3rd to June 8th, 2018 in Charleston, West Virginia
biblio-style: vancouver
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
    keep_tex: yes
    latex_engine: pdflatex
    template: template.tex
  html_document: default
blinded: 0
keywords:
- forensic science
- toolmark
- cross-correlation
- Mann-Whitney U statistic
- land engraved areas (LEAs)
- algorithm
- signatures 
- same-source problem
bibliography: bibliography
abstract: The same-source problem remains a major challenge in forensic toolmark and firearm examination. Technological advances in surface imaging allow measurements of 3D surfaces at previously unforeseen resolutions and enable digitized imaging. Here, we investigate the applicability of the Chumbley scoring method \citep{hadler}, developed for screwdriver markings, for same-source identification of bullet striae. We provide methods to identify parameters that minimize the error rates for matching of LEAs using Hamby datasets 44 and 252 measured by NIST and CSAFE. We suggest a remedial algorithm to alleviate the problem of failed tests in the method, increasing both the power of the test and reducing error rates. Type II error rates of the proposed method improve by more than one third (Type I error of 0.05) and are on average about 0.22. This puts the proposed method on similar footing as other single feature matching approaches in the literature. 
---
